---
title: "Data visualization"
output:
  html_document:
    df_print: paged
---

This is just an exercise solving notebook, trying to understand data visualization and carrying out the exercises as in the step  by step Data visualization from "R from data science" 


We shall be working with the "mpg" dataset. 

Among the variables in mpg we shall mainly work with these two variables, they are:

displ- a car’s engine size, in litres. </br>

hwy- a car’s fuel efficiency on the highway, in miles per gallon (mpg). 

```{r}
library(tidyverse)
```
```{r}
head(mpg)
```

Now we shall use ggplot to understand the relationship between "hwy" (fuel efficiency) and engine size "displ".

## 3.2.3 A graphing template


```{r}
ggplot(data=mpg)+ 
  geom_point(mapping=aes(x= displ, y=hwy))
```


We see that a car with a low fuel efficiency consumes more fuel than a car with a high fuel efficiency when they travel the same distanc, i.e. they encounter a negetive relationship, i.e. as the engine size increases the fuel efficiency of the car decreases. 

### Exercise 3.2.4: 

#### 1. Run ggplot(data = mpg). What do you see?

Gives a blank plot

```{r}
ggplot(data= mpg)
```



#### 2. How many rows are in mpg? How many columns?

The number of rows is 234
and columns 11

```{r}
dim(mpg) 
```
#### 3. What does the drv variable describe? Read the help for ?mpg to find out.

drv: The type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd </br>

#### 4. Make a scatterplot of hwy vs cyl.

Now creating another scatter plot between cyl and hwy

```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=cyl, y=hwy))
```


Here we see: 4. 

cyl: number of cylinders </br>
hwy: highway miles per gallon

We see that there is not a very defined relationship between the number of cylinders and the hwy, The graph suggests various ranges but at 7 cylinders it becomes stagnant. 

#### 5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

class: "type" of car </br>
drv: the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd

```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=class, y=drv))
```


This clearly shows what kind of car has what kind of drive train. However this plot is not useful as because, this does not show an particular relationship between the two. This does not show any correlation between the two variables. I.e. to say no positive or negetive relation is to be found as such. no proper regression line can be fit to the same. 

## 3.3 Aesthetic mappings

```{r}
ggplot(data=mpg)+ 
  geom_point(mapping=aes(x= displ, y=hwy, color=class))
```


### EXERCISE 3.3.1: 

#### 1. Whats wrong with the code?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```


When we use color inside the aesthetics, and as the color contains only one label as blue, so ir assigns the label blue to all the points, and the colour is given according to the first preference in the in its prefered list of colours for a discrete variable. hence the coloue used is "red", and the legend shows that all the points are marked "blue". 

#### 2: Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

dyspl, cty, hwy are continuous , rest are catagorical. 

#### 3: Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?


```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = cty ))
```



For color it creates a continuous shade, for shape it shows error as a continuous variable cannot be usd as there are only specifically 6 shapes that can be used at a time. 

For size it probably discretizes the continuous value but floor or ceiling function and assigs accordingly. 


```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = displ ))
```


#### 4: What happens if you map the same variable to multiple aesthetics?
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = displ, color= displ, alpha= displ))
```
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class, color= class, alpha= class, shape=class))
```



In this case it creates layers and applies all the aesthetics accoringly, like in the previous example we have used displ(a continuous variable) to size, colour and alpha and it worked good enough, however when we used class( a discrete catagorical variable) for multiple aesthetics, it gave the following error messages. 

#### 5: What does the stroke aesthetic do? What shapes does it work with?

```{r}
ggplot(mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(shape = 21, colour = "black", fill = "white", stroke = 2)

```


As we can see above the stokes gives the size of the border for the points, for which we use a particular shape, we can fill teh shape with "fill", increase the size using "size", chose colour using "color" and finally change the border using "stokes". 

It works with shape from 15 to 24. 

```{r}
ggplot(data=mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(shape = 24, colour = "black", fill = "white", stroke = 2)
```


#### 6: What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.

```{r}
ggplot(data=mpg)+ 
  geom_point(mapping = aes(x = displ, y = hwy, colour = displ<5))
```



As we see, this creates a binary response in case of colors. two colours are created according to the color palattee and at the same time, the legend shows a binary bifurcation depending on the size od display (this is because we gave 5, and display values are near to 5 ) 

##### doubt!!!!!  

```{r}
ggplot(data= mpg)+ 
  geom_point(mapping= aes(x= displ, y= hwy)) + 
  facet_wrap(~class, nrow=NULL)
```

```{r}
ggplot(data= mpg)+ 
  geom_point(mapping= aes(x= displ, y= hwy)) + 
  facet_wrap(~manufacturer, nrow=NULL)
```

```{r}
ggplot(data= mpg)+ 
  geom_point(mapping= aes(x= displ, y= hwy)) + 
  facet_grid(drv~cyl)
```

```{r}
ggplot(data= mpg)+ 
  geom_point(mapping= aes(x= displ, y= hwy)) + 
  facet_grid(. ~ cyl)
```

```{r}
ggplot(data=mpg)+geom_point(mapping = aes(x=displ, y= hwy))+
  facet_wrap(~cyl)
```

### 3.5.1 Exercises

#### 1: What happens if you facet on a continuous variable? 

```{r}
ggplot(data= mpg)+ 
  geom_point(mapping= aes(x= displ, y= hwy)) + 
  facet_wrap(~displ)
```
This does not seem to be very continuous output, lets look into a more continuous layout, lets use the iris dataset. 

```{r}
ggplot(data=iris)+ 
  geom_point(mapping=aes(x=Petal.Length, y=Petal.Width ))+
  facet_wrap(~Petal.Length)
```

As we can see in both the example there is one facet for each value, It creates a N*N grids depending on the total number of values. 


#### 2. What plots does the following code make? What does . do?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)
```
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

In the first plot: This divides the original layer of points in mapping displ against hwy, according to the drv attribute considering the levels of drv row wise. "." decides that the column division is undecided. Hence only row is considered. 

In the same way in the second plot only column is considerd according to the catagories of cyl. 

#### 4. Consider the first faceted plot in this section: What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

In colour aesthetics various colour are assigned to depict the points according to a particular attibute while in this the set of points are divided according to the catagories of the class and plotted seperately. 

With facetting is is easier ti classify the individual levels of an attribute, but colouring gives an overall view of the points. 

If we are interested in finding out how the individual catagory play a part in the overall data, the facetting is helpful. 

However for a large dataset if we are interested to find out the overall spread of the levels of an attribute wrt to all the points then colour aesthetic can be handy. 

HOpefully this answers all the points. 

#### 5. Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? Why doesn’t facet_grid() have nrow and ncol arguments?

rnow= number of rows we want to use, and ncol gives the number of columns. However it is also a way of determining how do we want our graph to look like (see the next two example). Other control options inculde as.table, dir etc. as.table tells the control to plot the levels of the attribute as in the table. Its default is TRUE. While dir gives the direction of the plot when we use nrows. The default direction is horizontal. 

```{r}
ggplot(data=iris)+ 
  geom_point(mapping=aes(x=Petal.Length, y=Petal.Width ))+
  facet_wrap(~Sepal.Length, nrow=3)
```


```{r}
ggplot(data=iris)+ 
  geom_point(mapping=aes(x=Petal.Length, y=Petal.Width ))+
  facet_wrap(~Sepal.Length, nrow=3, dir= "v") #same plot vertically directed
```

```{r}
ggplot(data=iris)+ 
  geom_point(mapping=aes(x=Petal.Length, y=Petal.Width ))+
  facet_wrap(~Sepal.Length, ncol=3, as.table = F) #as.table= F shows in the reverse order
```

#### 6. When using facet_grid() you should usually put the variable with more unique levels in the columns. Why?

```{r}
ggplot(data=iris)+ 
  geom_point(mapping=aes(x=Petal.Length, y=Petal.Width ))+
  facet_grid(Species~Sepal.Length)
```

```{r}
ggplot(data=iris)+ 
  geom_point(mapping=aes(x=Petal.Length, y=Petal.Width ))+
  facet_grid(Sepal.Length~Species)
```

The two plots above shows that, the space for marking the columns is more than the space allocated for marking the rows. So if more levels are present in the crow attribute then it will be difficult to properly identify the points, and hence will cause interpretation error.

Find the next set of exercises in the next file. 





