---
title: "PCA"
output:
  html_document:
    df_print: paged
---

Here we shall be applying PCA (Principal Component Analysis) to a data set, and stepwise show how the PCA is calculated in general, using inbuilt functions prcomp, princomp. 

Arguements taken by function prcomp: 

formula: formula with no respomse variable refering only to numeric value. </br>
data: an optional data frame containing the variables from the formula </br>

the default is:

x: a numeric matrix or data frame </br>
scale: a logical value indicating whether the variables should be scaled to have unit variance before the analysis takes place

Arguements takn by the function princomp: 

x: a numeric matrix or data frame </br>
cor: a logical value. If TRUE, the data will be centered and scaled before the analysis </br>
scores: a logical value. If TRUE, the coordinates on each principal component are calculated

The elements of the outputs returned by the functions prcomp() and princomp() includes :

```{r}
read.csv("required.csv", header = T)
```

Now we shall work with two different datasets. 


```{r}
data=read.table("pcadatawin.txt", header = T) # this is a imported data set.
head(data)
```

The dimension of the data set are as follows. 

```{r}
dim(data) 
```

Let us try to visualize the data. To obtain a 3D visualization we shall first </br>
1. Install a package called "devtools"
2. Then we use devtools::install_github("AckerDWM/gg3D") which downloads the gg3D directly from github repository AckerDWM. 

This package is useful in giving 3D plots showing 3 axes. The following example gives and empty plot.


```{r}

library("gg3D")

## An empty plot with 3 axes
qplot(x=0, y=0, z=0, geom="blank") + 
  theme_void() +
  axes_3D()

```

This will give the necessary 3D representation the data points for iris dataset, and the imported data set respectively.  

```{r}
data(iris)
ggplot(iris, aes(x=Petal.Width, y=Sepal.Width, z=Petal.Length, color=Species)) + 
  theme_void() +
  axes_3D() +
  stat_3D()

```

```{r}
ggplot(data, aes(x=x, y=y, z=z)) + 
  theme_void() +
  axes_3D() +
  stat_3D()

```

The last three plots give a visualization of the data, The first plot shows that the data is spread in the x-y plane and partly in the z axis, the second plot for the other set of data shows that the points are spread mostly in the x-y plane. This gives a intuition about what can be the Principal components. Next we shall apply "prcomp()" to find out the PCs and the the scree plot for both the set of datas. 

```{r}
library(factoextra)
pc=prcomp(data, scale= FALSE)
pc
```
```{r}
pc1=princomp(data, scores = TRUE) #using princomp
pc1

```
```{r}
names(pc1)
```


```{r}
names(pc)

```

We see that both prcomp and princomp gives nearly the same values. x in prcomp are the scores in primcomp. 

```{r}
pc_plotdata=data.frame(v=pc$sdev)^2
ggplot(data=pc_plotdata)+
  geom_col(mapping = aes(x=c(1,2,3), y=v)) #same as bar taking stat= "identity by default
```

```{r}
fviz_eig(pc, addlabels = TRUE)
```

```{r}
fviz_screeplot(pc, choice="eigenvalue", 
               barfill= "blue", addlabels= TRUE)
```
Now working with IRIS dataset. 


```{r}
iris1=iris[-5]
pc=prcomp(iris1, scale= FALSE)
pc

```

```{r}
pc1=princomp(iris1, scores = TRUE) #using princomp
pc1

```

```{r}
fviz_screeplot(pc1, choice="eigenvalue", 
               barfill= "blue", addlabels= TRUE)

```

